\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-29 paper.qmd
first two paragraphs of background

<!-- 
Check-list before submission
* Is it all American spelling
* Spelling checked generally
* Code all runs given fresh workspace
* Code has a readme, explaining how the paper results are reproduced
-->

@@

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional representation of high-dimensional data. Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008], uniform manifold approximation and projection (UMAP) [@Leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@article03], large-scale dimensionality reduction Using triplets (TriMAP) [@article02], and pairwise controlled manifold approximation (PaCMAP) [@Yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
# 2D layouts from different NLDR techniques and different hyperparameter choices applied for the PBMC3k dataset: (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (d) UMAP (n_neighbors = 15, min_dist = 0.03), (e) UMAP (n_neighbors = 84, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (h) tSNE (perplexity = 92), (i) PHATE (knn = 2), (j) PHATE (knn = 5), (k) PHATE (knn = 10), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (m) TriMAP (n_inliers = 30, n_outliers = 4, n_random = 10), (n) TriMAP (n_inliers = 5, n_outliers = 2, n_random = 2), (p) PaCMAP (n_neighbors = 10, init = pca, MN_ratio = 0.5, FP_ratio = 2), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5), (r) PaCMAP (n_neighbors = 5, init = pca, MN_ratio = 0.1, FP_ratio = 1), and (s) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.1, FP_ratio = 1). Is there a best representation of the original data or are they all providing  equivalent information?
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

@@

  - Create a representation of the model
  - Algorithm in 2D
      - Parameters
      - Tuning
  - Showing model in high-d
  - What is learned about simulated examples

>>>>

<!-- 
Check-list before submission
* Is it all American spelling
* Spelling checked generally
* Code all runs given fresh workspace
* Code has a readme, explaining how the paper results are reproduced
* Re-write abstract
-->

@@

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional ($k-D$) representation of high-dimensional ($p-D$) data. Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008], uniform manifold approximation and projection (UMAP) [@Leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@article03], large-scale dimensionality reduction Using triplets (TriMAP) [@article02], and pairwise controlled manifold approximation (PaCMAP) [@Yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

<!-- - Connection between NLDR and MDS-->
Historically, $k-D$ representations of $p-D$ data have been computed using multidimensional Scaling (MDS), which includes principal components analysis as a special case.  The $k-D$ representation can be considered to be a layout of points in $k-D$ produced by an embedding procedure that maps the data from $p-D$. In MDS, the $k-D$ layout is constructed by minimizing a stress function that differences distances between points in $p-D$ with potential distances between points in $k-D$. Various formulations of the stress function result in non-metric scaling [@article62] and isomap [@article63]. Challenges in working with high-dimensional data, including visualization, are outlined in @Johnstone2009.

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in $p-D$. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the $k-D$ minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@article64] spreading to capture geometric shapes, that include both global and local structure.

- Tours
- Model-in-the-data-space: how can we represent the model, eg plane for PCA, grid of values for classification boundaries.

@@

- Overview: Generate a form that maps the model, that is the interpoint distances. What is the model?
- Notation
- Create a representation of the model 
    - using hex-binning in 2D, 
    - parameters, 
    - tuning, 
    - pre-processing
- How does this map to the representation in high-d
    - Centroids, 
    - Edges
- Measuring fit
    - Fitted values
    - Error calculation
- What is learned about simulated examples
    - Interesting organisation of points in UMAP
    -

### 2024-03-29 paper.tex
first two paragraphs of background

Non-linear dimension reduction (NLDR) is popular for making a convenient
low-dimensional representation of high-dimensional data. Recently
developed methods include t-distributed stochastic neighbor embedding
(tSNE) \citep{Laurens2008}, uniform manifold approximation and
projection (UMAP) \citep{Leland2018}, potential of heat-diffusion for
affinity-based trajectory embedding (PHATE) algorithm \citep{article03},
large-scale dimensionality reduction Using triplets (TriMAP)
\citep{article02}, and pairwise controlled manifold approximation
(PaCMAP) \citep{Yingfan2021}. However, the representation generated can
vary dramatically from method to method, and with different choices of
parameters or random seeds made using the same method
(Figure~\ref{fig-NLDR-variety}). The dilemma for the analyst is then,
\textbf{which representation to use}. The choice might result in
different procedures used in the downstream analysis, or different
inferential conclusions. The research described here provides new visual
tools to aid with this decision.

@@

@@

\begin{itemize}
\tightlist
\item
  Create a representation of the model
\item
  Algorithm in 2D

  \begin{itemize}
  \tightlist
  \item
    Parameters
  \item
    Tuning
  \end{itemize}
\item
  Showing model in high-d
\item
  What is learned about simulated examples
\end{itemize}

>>>>

Non-linear dimension reduction (NLDR) is popular for making a convenient
low-dimensional (\(k-D\)) representation of high-dimensional (\(p-D\))
data. Recently developed methods include t-distributed stochastic
neighbor embedding (tSNE) \citep{Laurens2008}, uniform manifold
approximation and projection (UMAP) \citep{Leland2018}, potential of
heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm
\citep{article03}, large-scale dimensionality reduction Using triplets
(TriMAP) \citep{article02}, and pairwise controlled manifold
approximation (PaCMAP) \citep{Yingfan2021}. However, the representation
generated can vary dramatically from method to method, and with
different choices of parameters or random seeds made using the same
method (Figure~\ref{fig-NLDR-variety}). The dilemma for the analyst is
then, \textbf{which representation to use}. The choice might result in
different procedures used in the downstream analysis, or different
inferential conclusions. The research described here provides new visual
tools to aid with this decision.

@@

Historically, \(k-D\) representations of \(p-D\) data have been computed
using multidimensional Scaling (MDS), which includes principal
components analysis as a special case. The \(k-D\) representation can be
considered to be a layout of points in \(k-D\) produced by an embedding
procedure that maps the data from \(p-D\). In MDS, the \(k-D\) layout is
constructed by minimizing a stress function that differences distances
between points in \(p-D\) with potential distances between points in
\(k-D\). Various formulations of the stress function result in
non-metric scaling \citep{article62} and isomap \citep{article63}.
Challenges in working with high-dimensional data, including
visualization, are outlined in \citet{Johnstone2009}.

Many new methods for NLDR have emerged in recent years, all designed to
better capture specific structures potentially existing in \(p-D\). Here
we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP
and PaCMAP. tNSE and UMAP can be considered to produce the \(k-D\)
minimizing the divergence between two distributions, where the
distributions are modeling the inter-point distances. PHATE, TriMAP and
PaCMAP are examples of diffusion processes \citep{article64} spreading
to capture geometric shapes, that include both global and local
structure.

\begin{itemize}
\tightlist
\item
  Tours
\item
  Model-in-the-data-space: how can we represent the model, eg plane for
  PCA, grid of values for classification boundaries.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Overview: Generate a form that maps the model, that is the interpoint
  distances. What is the model?
\item
  Notation
\item
  Create a representation of the model

  \begin{itemize}
  \tightlist
  \item
    using hex-binning in 2D,
  \item
    parameters,
  \item
    tuning,
  \item
    pre-processing
  \end{itemize}
\item
  How does this map to the representation in high-d

  \begin{itemize}
  \tightlist
  \item
    Centroids,
  \item
    Edges
  \end{itemize}
\item
  Measuring fit

  \begin{itemize}
  \tightlist
  \item
    Fitted values
  \item
    Error calculation
  \end{itemize}
\item
  What is learned about simulated examples

  \begin{itemize}
  \tightlist
  \item
    Interesting organisation of points in UMAP
  \item
  \end{itemize}
\end{itemize}

### 2024-03-29 paper_original.tex
first two paragraphs of background

To investigate which is the best representation to visualize the
original data or all NLDR methods provide equivalent information, we
visualize the model-in-data space. Models from all NLDR methods show
five well-separated clusters (see Figure~\ref{fig-gau1-sc},
Figure~\ref{fig-gau2-sc}, Figure~\ref{fig-gau3-sc},
Figure~\ref{fig-gau4-sc}, and Figure~\ref{fig-gau5-sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure. tSNE, UMAP, and PHATE display clusters
with varying densities, indicating their ability to capture
within-cluster variation (see Figure~\ref{fig-gau1-sc},
Figure~\ref{fig-gau2-sc}, and Figure~\ref{fig-gau3-sc}). On the other
hand, both TriMAP and PaCMAP show clusters with flat surfaces,
suggesting a failure to capture within-cluster variation (see
Figure~\ref{fig-gau4-sc} and Figure~\ref{fig-gau5-sc}). Therefore,
TriMAP and PaCMAP do not capture the local structure as effectively as
other methods.

@@

\subcaption{\label{fig-gau1-sc1}}

@@

\subcaption{\label{fig-gau1-sc2}}

@@

\subcaption{\label{fig-gau1-sc3}}

@@

\caption{\label{fig-gau1-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/oQxEb4wRdHI}).}

@@

\subcaption{\label{fig-gau2-sc1}}

@@

\subcaption{\label{fig-gau2-sc2}}

@@

\subcaption{\label{fig-gau2-sc3}}

@@

\caption{\label{fig-gau2-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/JW49csPpDx4}).}

@@

\subcaption{\label{fig-gau3-sc1}}

@@

\subcaption{\label{fig-gau3-sc2}}

@@

\subcaption{\label{fig-gau3-sc3}}

@@

\caption{\label{fig-gau3-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/xqhRODt8Z-o}).}

@@

\subcaption{\label{fig-gau4-sc1}}

@@

\subcaption{\label{fig-gau4-sc2}}

@@

\subcaption{\label{fig-gau4-sc3}}

@@

\caption{\label{fig-gau4-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/X-uHSB4VoZ0}).}

@@

\subcaption{\label{fig-gau5-sc1}}

@@

\subcaption{\label{fig-gau5-sc2}}

@@

\subcaption{\label{fig-gau5-sc3}}

@@

\caption{\label{fig-gau5-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/UxvU8HQd7JM}).}

@@

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the high-D data
(model-in-data space). The figures in Figure~\ref{fig-pbmc1-sc} show
three well-separated clusters, indicating that the suggested UMAP
representation preserves the global structure (see
Figure~\ref{fig-umapauthor}). However, as shown in
Figure~\ref{fig-pbmc1-sc}, these clusters are close to each other in
high-D space. Also, nonlinear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1-sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umapauthor}) does not accurately preserve the local
structure of the PBMC3k dataset.

@@

\subcaption{\label{fig-pbmc1-sc1}}

@@

\subcaption{\label{fig-pbmc1-sc2}}

@@

\subcaption{\label{fig-pbmc1-sc3}}

@@

\caption{\label{fig-pbmc1-sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/Gnhh4hfsbbg}).}

@@

Next, the model is fitted for tSNE, and the resulting model is
visualized in the data space. The model shows some quirks, as shown in
Figure~\ref{fig-pbmc2-sc}. The large cluster is divided into three
subclusters, which are in close to each other.

@@

\subcaption{\label{fig-pbmc2-sc4}}

@@

\subcaption{\label{fig-pbmc2-sc5}}

@@

\subcaption{\label{fig-pbmc2-sc6}}

@@

\caption{\label{fig-pbmc2-sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/R0EroJcGbas}).}

@@

According to Figure~\ref{fig-mnist1-sc1}, the nonlinear continuity
structure observed in the 2D representation of PaCMAP is also visible
when visualizing the model overlaid on the data space. This provides
evidence that the model accurately captures the structure of the high-D
data. Also, the model has a twisted pattern within the nonlinear
structure, which is an addition pattern of data that cannot be seen in
the 2D representation (see Figure~\ref{fig-mnist1-sc2}).

In order to visually assess the preservation of local structure, it's
important to consider the edges in the model. In a model that
effectively preserves local structure, there shouldn't be any long edges
in high-D space. However, as shown in Figure~\ref{fig-mnist1-sc3}, some
long edges exist in high-D space that are not identified as long edges
in the 2D representation.

@@

\subcaption{\label{fig-mnist1-sc1}}

@@

\subcaption{\label{fig-mnist1-sc2}}

@@

\subcaption{\label{fig-mnist1-sc3}}

@@

\caption{\label{fig-mnist1-sc}Screen shots of the \textbf{langevitour}
of the MNIST digit 1 data set, shows the model-in-data space, a video of
the tour animation is available at
(\url{https://youtu.be/zq2GM9qvUNA}).}

>>>>

To investigate which is the best representation to visualize the
original data or all NLDR methods provide equivalent information, we
visualize the model-in-data space. Models from all NLDR methods show
five well-separated clusters (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, Figure~\ref{fig-gau3_sc},
Figure~\ref{fig-gau4_sc}, and Figure~\ref{fig-gau5_sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure. tSNE, UMAP, and PHATE display clusters
with varying densities, indicating their ability to capture
within-cluster variation (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, and Figure~\ref{fig-gau3_sc}). On the other
hand, both TriMAP and PaCMAP show clusters with flat surfaces,
suggesting a failure to capture within-cluster variation (see
Figure~\ref{fig-gau4_sc} and Figure~\ref{fig-gau5_sc}). Therefore,
TriMAP and PaCMAP do not capture the local structure as effectively as
other methods.

@@

\subcaption{\label{fig-gau1\_sc1}}

@@

\subcaption{\label{fig-gau1\_sc2}}

@@

\subcaption{\label{fig-gau1\_sc3}}

@@

\caption{\label{fig-gau1_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/oQxEb4wRdHI}).}

@@

\subcaption{\label{fig-gau2\_sc1}}

@@

\subcaption{\label{fig-gau2\_sc2}}

@@

\subcaption{\label{fig-gau2\_sc3}}

@@

\caption{\label{fig-gau2_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/JW49csPpDx4}).}

@@

\subcaption{\label{fig-gau3\_sc1}}

@@

\subcaption{\label{fig-gau3\_sc2}}

@@

\subcaption{\label{fig-gau3\_sc3}}

@@

\caption{\label{fig-gau3_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/xqhRODt8Z-o}).}

@@

\subcaption{\label{fig-gau4\_sc1}}

@@

\subcaption{\label{fig-gau4\_sc2}}

@@

\subcaption{\label{fig-gau4\_sc3}}

@@

\caption{\label{fig-gau4_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/X-uHSB4VoZ0}).}

@@

\subcaption{\label{fig-gau5\_sc1}}

@@

\subcaption{\label{fig-gau5\_sc2}}

@@

\subcaption{\label{fig-gau5\_sc3}}

@@

\caption{\label{fig-gau5_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/UxvU8HQd7JM}).}

@@

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the high-D data
(model-in-data space). The figures in Figure~\ref{fig-pbmc1_sc} show
three well-separated clusters, indicating that the suggested UMAP
representation preserves the global structure (see
Figure~\ref{fig-umapauthor}). However, as shown in
Figure~\ref{fig-pbmc1_sc}, these clusters are close to each other in
high-D space. Also, nonlinear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1_sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umapauthor}) does not accurately preserve the local
structure of the PBMC3k dataset.

@@

\subcaption{\label{fig-pbmc1\_sc1}}

@@

\subcaption{\label{fig-pbmc1\_sc2}}

@@

\subcaption{\label{fig-pbmc1\_sc3}}

@@

\caption{\label{fig-pbmc1_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/Gnhh4hfsbbg}).}

@@

Next, the model is fitted for tSNE, and the resulting model is
visualized in the data space. The model shows some quirks, as shown in
Figure~\ref{fig-pbmc2_sc}. The large cluster is divided into three
subclusters, which are in close to each other.

@@

\subcaption{\label{fig-pbmc2\_sc4}}

@@

\subcaption{\label{fig-pbmc2\_sc5}}

@@

\subcaption{\label{fig-pbmc2\_sc6}}

@@

\caption{\label{fig-pbmc2_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/R0EroJcGbas}).}

@@

According to Figure~\ref{fig-mnist1_sc1}, the nonlinear continuity
structure observed in the 2D representation of PaCMAP is also visible
when visualizing the model overlaid on the data space. This provides
evidence that the model accurately captures the structure of the high-D
data. Also, the model has a twisted pattern within the nonlinear
structure, which is an addition pattern of data that cannot be seen in
the 2D representation (see Figure~\ref{fig-mnist1_sc2}).

In order to visually assess the preservation of local structure, it's
important to consider the edges in the model. In a model that
effectively preserves local structure, there shouldn't be any long edges
in high-D space. However, as shown in Figure~\ref{fig-mnist1_sc3}, some
long edges exist in high-D space that are not identified as long edges
in the 2D representation.

@@

\subcaption{\label{fig-mnist1\_sc1}}

@@

\subcaption{\label{fig-mnist1\_sc2}}

@@

\subcaption{\label{fig-mnist1\_sc3}}

@@

\caption{\label{fig-mnist1_sc}Screen shots of the \textbf{langevitour}
of the MNIST digit 1 data set, shows the model-in-data space, a video of
the tour animation is available at
(\url{https://youtu.be/zq2GM9qvUNA}).}
